Association between trauma severity models and opportunities for improvement:

### 2022-12-15 manuscript.Rmd
Comment on project paper and table function

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  pdf_document: default
  html_document:
    df_print: paged
    number_sections: true
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide and accounts for approximately 9% of global deaths annually [@globalburden]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. The quality of trauma care is therefore a globally relevant issue with significance for a great number of people. Improvements in trauma outcomes benefit both the individual and also society at large by reducing the societal and healthcare costs of disability. The significance of the consequences and burden of trauma has been recognised by the UN, who in their Sustainable Development Goals included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg]. The importance of improving trauma care named as one of the strategies in achieving this goal. The goal was not met by 2020 [@sdg20], which is another signal that there is significant progress to be made in trauma care.

Efforts to advance the state of trauma care have been ongoing for many years. Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi]. QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference. These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case. One method of improving the quality of M&M conferences is by improving case selection. This is usually achieved by using audit filters that automatically flag cases for review. One approach is the use of trauma severity models as audit filters. Trauma severity models are algorithms that evaluate trauma severity in quantitative terms. The effectiveness of these models in case selection is understudied and their potential as audit filters is the focus of this paper.

@@

The significance of M&M conferences in trauma QI is a consequence of the unique challenges inherent to trauma care. Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines. Outcomes in severe trauma are dependent on the quality of care received [@ustqi] and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care. This can only be achieved through peer review and consequently, M&M conferences are the cornerstone of trauma QI programs.

As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care. Regular conferences allow QI programs to identify patterns of errors and provide a basis for systemic quality improvement. To maximise effectiveness, conferences should be attended by all involved specialities and have a structured format. The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines]. Despite theserecommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

Karolinska University Hospital is located in Solna, Sweden. The trauma centre at Karolinska University Hospital is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats] and it is the primary trauma centre for the region of Stockholm. Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats]. 
The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines. All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre. Data is collected from the patient's electronic medical records. The trauma registry contains data on patients from 2013 to 2021. <!-- Based on the dates I see in the Swetrau dataset --> 
<!--JA: Det har samlats ännu äldre data i olika former. Skulle bara skippa årtalen kring traumaregistret. KS har kört någon form av MoM sedan 2012. Se: https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/24625137/ -->

@@

```{r, echo=FALSE, out.width = "400px"}
knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
```

@@

The algorithm for PS includes the patient's age, gender, GCS and ISS.

@@

The code for data analysis was written using simulated data to reduce the risk of research bias. Real-world data was only used once the data analysis model was completed and found to work correctly on simulated data.

@@

R, a programming language for statistical computing, was used for data collection and analysis. Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR. The CI for the OR was computed using the Wald estimator. Receiver operator characteristic (ROC) curves were used to assess diagnostic performance. The area under the curve (AUC) was calculated to compare the models with one another. The Integrated Calibration Index (ICI) was used to measure calibration. Accuracy was plotted and is reported as the highest value for accuracy in the plot together with the associated cutoff. Results werepresented with a confidence interval of 95%. A p-value of <0.05 was considered significant.

@@

Participants
------------
A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period. Of these, `r exclusion[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion. `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`). Figure 2 shows the inclusion process for eligible participants.

@@

@@

Main results
------------
<!--Figure X??? -->
The fitted curves for the logistic regression models of each score are shown in figure x. The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.

```{r}
knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
```
<!--Figure X??? -->
The receiver operating curves for all scores are plotted in figure x. The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.

```{r}
knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
```
<!--Figure X??? -->
The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively. The accuracy at all cutoffs is plotted for all scores in figure x. The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.

```{r}
knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
```
<!--Table X? -->
The differences between the scores for each result are shown in table x.

@@

Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions. Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful. Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed. Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity. They are much less useful when applied to individuals, and were not intended for that use.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pepe2019; @trentzsch2004], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

@@

This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, opportunities for improvement, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences. We showed that, while low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI. The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.

>>>>

---
## To use of cross-references and avoid having to manually number figures and tables 
## you need to switch to the output format bookdown::pdf_document2 and 
## bookdown::word_document2, see https://bookdown.org/yihui/rmarkdown-cookbook/cross-ref.html
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  pdf_document: default
  html_document:
    df_print: paged
    number_sections: true
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pts
linestretch: 1
always_allow_html: yes
editor_options: 
   chunk_output_type: inline
---

@@

Introduction
============
<!-- I think that it often makes sense to keep paragraphs short and focus on one key message per paragraph. In this first paragraph you present multiple important points, the burden of fatal trauma, the burden of non-fatal trauma, and the importance of improving trauma care. Each of these messages can have their own, shorter paragraphs, which in my opinion makes any text easier to digest.-->
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide and accounts for approximately 9% of global deaths annually [@globalburden]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. The quality of trauma care is therefore a globally relevant issue with significance for a great number of people. Improvements in trauma outcomes benefit both the individual and also society at large by reducing the societal and healthcare costs of disability. The significance of the consequences and burden of trauma has been recognised by the UN, who in their Sustainable Development Goals included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg]. The importance of improving trauma care named as one of the strategies in achieving this goal. The goal was not met by 2020 [@sdg20], which is another signal that there is significant progress to be made in trauma care.

<!-- See my comment above. You introduce several important concepts here, and one that is key to your paper, the use of trauma severity models for case selection. By introducing it late in the paragraph, you effectively hide it, and its importance is kind of lost on the reader. I suggest that you split this paragraph in two, and devote the second one to the use of trauma severity models-->
Efforts to advance the state of trauma care have been ongoing for many years. Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi]. QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference. These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case. One method of improving the quality of M&M conferences is by improving case selection. This is usually achieved by using audit filters that automatically flag cases for review. One approach is the use of trauma severity models as audit filters. Trauma severity models are algorithms that evaluate trauma severity in quantitative terms. The effectiveness of these models in case selection is understudied and their potential as audit filters is the focus of this paper. <!-- This last sentence is perhaps the most important sentence in the introduction, but that is not very clear from how it is presented. See if you can highlight it more, perhaps by having it as the first sentence in a separate paragraph.-->

@@

The significance of M&M conferences in trauma QI is a consequence of the unique challenges inherent to trauma care. Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines. Outcomes in severe trauma are dependent on the quality of care received [@ustqi]<!-- It's uncommon to introduce a citation in the middle of a sentence. See if you can rephrase the sentence to allow it to be placed last, or divide the sentence somehow, perhaps with a ",", and place the citation just before that.-->  and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care. This can only be achieved through peer review and consequently, M&M conferences are the cornerstone of trauma QI programs.

As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions<!-- It's okay to be more direct: "The purpose of M&M conferences is to review selected patient cases, establish if there were opportunities for improvement, and implement corrective actions"-->"" . The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care. Regular conferences allow QI programs to identify patterns of errors and provide a basis for systemic quality improvement. To maximise effectiveness, conferences should be attended by all involved specialities and have a structured format. The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines]. Despite these recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

Karolinska University Hospital is located in Solna, Sweden. The trauma centre at Karolinska University Hospital is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats] and it is the primary trauma centre for the region of Stockholm. Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats].

The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines. All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre. Data is collected from the patient's electronic medical records. The trauma registry contains data on patients from 2013 to 2021. <!-- Based on the dates I see in the Swetrau dataset --> 
<!--JA: Det har samlats ännu äldre data i olika former. Skulle bara skippa årtalen kring traumaregistret. KS har kört någon form av MoM sedan 2012. Se: https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/24625137/ -->

@@

```{r, echo=FALSE, out.width = "400px"}
knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
```
<!-- In the figure it looks like all cases are reviewed by a second nurse, which is not true. Most cases are not reviewed by a second nurse, because either none of the audit filters are positive or the first nurse decides that there were no issues. So I would add an arrow to the right of the box "Audit filters and primary review..." stating that some cases are excluded at this stage. Then I would add a second arrow to the right of "Secondary review..." stating that additional cases are excluded at this step, if no issues are found. So it's only a small subset of all cases that are reviewed at a M&M conference-->

@@

The algorithm for PS includes the patient's age, gender, GCS and ISS. <!-- This is the older PS, right? for the paper I would mention that you used that version because the data necessary for the new version is not included in the registry -->

@@

The code for data analysis was written using simulated data to reduce the risk of researcher bias. Real-world data was only used once the data analysis model was completed and found to work correctly on simulated data.

@@

R, a programming language for statistical computing, was used for data collection and analysis<!-- Add citation-->. Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR. The CI for the OR was computed using the Wald estimator. Receiver operator characteristic (ROC) curves were used to assess diagnostic performance. The area under the curve (AUC) was calculated to compare the models with one another. The Integrated Calibration Index (ICI) was used to measure calibration. Accuracy was plotted and is reported as the highest value for accuracy in the plot together with the associated cutoff. Results are presented with a confidence interval of 95%. A p-value of <0.05 was considered significant.

@@

Participants
------------
A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period. Of these, `r exclusion[2,2]` were reviewed at an M&M conference<!-- Not really. Only a small subset of cases were reviewed at a conference, but all were part of the review process--> and therefore became eligible for study inclusion. `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`). Figure 2 shows the inclusion process for eligible participants.

@@

<!-- See my comment in your table.R, but the above could be: -->
```{r incomplete-table, echo=FALSE}
knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data")
```

<!-- And you can then reference this table from anywhere in the text with \@ref(tab:incomplete-table), which would take care of the numbering for you. -->

@@

Main results
------------
<!--Figure X??? -->
<!-- You can also cross reference figures like this: \@ref(fig:fitted-glm) -->
The fitted curves for the logistic regression models of each score are shown in figure x. The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.

```{r fitted-glm, fig.cap = "Fitted curves for the logistic regression models of each score"}
knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
## Make sure you label the axes
```
<!--Figure X??? -->
The receiver operating curves for all scores are plotted in figure x. The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.

```{r}
knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
## Write Sensitivity and Specificity, with capital S
```
<!--Figure X??? -->
The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively. The accuracy at all cutoffs is plotted for all scores in figure x. The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.

```{r}
knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
```
<!--Table X? -->
The differences between the scores for each result are shown in table x. <!-- Elaborate on what differences are shown in the table. You are actually showing the differences in performance measures, right? So here you can also say something about what models performed best in terms of what measure-->

@@

Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions. Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful. Audit filters that don't<!-- Avoid contractions--> improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed. Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity. They are much less useful when applied to individuals, and were not intended for that use.

@@

### Health equity
<!-- Is this section requested by KI?-->
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pepe2019; @trentzsch2004], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

@@

This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, opportunities for improvement, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences<!-- No need to state your aim again here in the conclusion, unless it's specifically asked for by KI.-->. We showed that, while low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI. The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.
